---
title: "abundance_calculations"
author: "Maddy Enda"
format: html
editor: visual
---

## Call libraries

```{r}
# Load in the packages for this data analysis
librarian::shelf(tidyverse, dplyr, terra, tmap, sf, here, stars, janitor, units, showtext, knitr, kableExtra)
# Adding in our text
font_add_google("Montserrat", "montserrat")
showtext_auto()
```

## Assign palette

```{r}
# Create palette
biota_pal <- c("seagreen", "#A3AD62FF", "#D0D3A2", "#798234FF", "#F0C6C3FF", "#DF91A3FF", "#D46780FF" )
```

## Statewide Analysis:

### Find the total area of state waters using the state water boundary

```{r}
# Read in the study regions data
state_boundaries <-  sf::st_read("MLPA_Study_Regions")
```

```{r}
# Transform to the UTM 10N crs for California
state_boundary_transform <- st_transform(state_boundaries, crs= 32610)
statewater_area <- sum(st_area(state_boundary_transform) / 10000)
```

```{r}
# Calculate area in hectares
sr_boundaries <- state_boundary_transform |>
  mutate(area_ha = as.numeric(st_area(geometry)) / 10000)
```

```{r}
scsr_area <- 795004.7
ccsr_area <- 296913.5
nccsr_area <- 197503.2
ncsr_area <- 266790.1
```

### Find area in hectares by biota category

```{r}
# Read in the prefiltered biota data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
ca_biota <- readRDS(file.path(rds.dir, "biota.rds"))

# Remove the unclassified observations
ca_biota_clean <- ca_biota %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop the geometries to speed up run time
ca_biota_simple <- ca_biota_clean %>%
  st_drop_geometry()

# Create a column that has the total area of socal biota
ca_biota_simple <- ca_biota_simple %>%
  mutate(total_area = sum(area_ha))
```

```{r}
statewater_area
```

```{r}
unique(ca_biota_simple$total_area)
```

### Get absolute abundance for biota

```{r}
# Create a sum of area column in mpa_transform
ca_biota_simple_2 <- ca_biota_simple %>%
  group_by(cmecs_bc_category) %>%  
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area / statewater_area) * 100)
```

```{r}
 # Create table
kable(ca_biota_simple_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total Statewaters")) %>%
  kable_material(c("striped", "hover"))
```

### Repeating our calculations within the MPA system

```{r}
# Load in the MPA boundary data
MPA_boundary <- sf::st_read("data/California_Marine_Protected_Areas_[ds582]")

# Filter to Point Lobos locations
mpa <- MPA_boundary %>%
   clean_names()
```

```{r}
# Reproject to UTM 10
mpa_transform <- st_transform(mpa, crs= 32610)
mpa_area <-  sum(st_area(mpa_transform) / 10000)
```

```{r}
# Join the data sets together
ca_biota_mpa <- st_intersection(ca_biota_clean, mpa_transform)
```

```{r}
ca_biota_mpa$biota_area_ha = (st_area(ca_biota_mpa) / 10000)
```

```{r}
# Remove geometries to run faster
ca_biota_mpa_simple <- ca_biota_mpa %>%
  st_drop_geometry()
```

```{r}
# Create a sum of area column in mpa_transform
ca_biota_mpa_2 <- ca_biota_mpa_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_cat_ha = sum(biota_area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_cat_ha / 262882) * 100)
```

```{r}
 # Create table
kable(ca_biota_mpa_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total MPA Area")) %>%
  kable_material(c("striped", "hover"))
```

## Abundance in SCSR:

### Region wide:

```{r}
# Make regional boundaries
scsr <- state_boundary_transform %>%
  filter(Study_Regi == "SCSR")

ccsr <- state_boundary_transform %>%
  filter(Study_Regi == "CCSR")

nccsr <- state_boundary_transform %>%
  filter(Study_Regi == "NCCSR")

ncsr <- state_boundary_transform %>%
  filter(Study_Regi == "NCSR")
```

```{r}
# Make valid the scsr geometry
scsr <- scsr %>% 
  st_make_valid()

# Find all biota in scsr region
scsr_biota <- st_intersection(scsr,ca_biota_clean)
```

```{r}
# Remove the unclassified observations
scsr_biota_clean <- scsr_biota %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
scsr_biota_simple <- scsr_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
biota_scsr <- scsr_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /scsr_area) * 100)
```

```{r}
 # Create table
kable(biota_scsr, col.names = c("Biota Category", "Area in hectares", "Percent of Total SCSR Area")) %>%
  kable_material(c("striped", "hover"))
```

### In SCSR MPAs:

```{r}
# Read in the prefiltered biota data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
socal_biota <- readRDS(file.path(rds.dir, "scsr_biota.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
socal_biota_transform <- st_transform(socal_biota, crs = 32610)

# Calculate area in hectares
socal_biota_transform$socal_ha <- (st_area(socal_biota_transform) / 10000)

# Remove the unclassified observations
socal_biota_clean <- socal_biota_transform %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
socal_biota_simple <- socal_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
socal_biota_simple_2 <- socal_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(socal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /scsr_area ) * 100)
```


```{r}
 # Create table
kable(socal_biota_simple_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total South Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

## Abundance in CCSR:

### In CCSR region

```{r}
# Make valid the scsr geometry
ccsr <- ccsr %>% 
  st_make_valid()

# Find all biota in scsr region
ccsr_biota <- st_intersection(ccsr,ca_biota_clean)
```

```{r}
# Remove the unclassified observations
ccsr_biota_clean <- ccsr_biota %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
ccsr_biota_simple <- ccsr_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
biota_ccsr <- ccsr_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /ccsr_area) * 100)
```

```{r}
 # Create table
kable(biota_ccsr, col.names = c("Biota Category", "Area in hectares", "Percent of Total Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In CCSR MPAs:

```{r}
# Read in the prefiltered biota data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
central_biota <- readRDS(file.path(rds.dir, "ccsr_biota.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
cencal_biota_transform <- st_transform(central_biota, crs = 32610)

# Calculate area in hectares
cencal_biota_transform$cencal_ha <- (st_area(cencal_biota_transform) / 10000)

# Remove the unclassified observations
cencal_biota_clean <- cencal_biota_transform %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
cencal_biota_simple <- cencal_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
cencal_biota_simple_2 <- cencal_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(cencal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /ccsr_area ) * 100)
```

```{r}
 # Create table
kable(cencal_biota_simple_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total Central Coast Region MPAS")) %>%
  kable_material(c("striped", "hover"))
```

## Abundance in NCCSR:

### In NCCSR Region

```{r}
# Make valid the scsr geometry
nccsr <- nccsr %>% 
  st_make_valid()

# Find all biota in scsr region
nccsr_biota <- st_intersection(nccsr,ca_biota_clean)
```

```{r}
# Remove the unclassified observations
nccsr_biota_clean <- nccsr_biota %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
nccsr_biota_simple <- nccsr_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
biota_nccsr <- nccsr_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /nccsr_area) * 100)
```

```{r}
 # Create table
kable(biota_nccsr, col.names = c("Biota Category", "Area in hectares", "Percent of Total North Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In NCCSR MPAs

```{r}
# Read in the prefiltered biota data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
nccsr_biota <- readRDS(file.path(rds.dir, "nccsr_biota.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
nccsr_biota_transform <- st_transform(nccsr_biota, crs = 32610)

# Calculate area in hectares
nccsr_biota_transform$nccsr_ha <- (st_area(nccsr_biota_transform) / 10000)

# Remove the unclassified observations
nccsr_biota_clean <- nccsr_biota_transform %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
nccsr_biota_simple <- nccsr_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column 
nccsr_biota_simple_2 <- nccsr_biota_simple %>%
  group_by(cmecs_bc_category) %>%  
  summarise(
    biota_area = sum(nccsr_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /nccsr_area ) * 100)
```

```{r}
 # Create table
kable(nccsr_biota_simple_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total North Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

## Abundance in NCSR:

### In the NCSR Region:

```{r}
# Make valid the scsr geometry
ncsr <- ncsr %>% 
  st_make_valid()

# Find all biota in scsr region
ncsr_biota <- st_intersection(ncsr,ca_biota_clean)
```

```{r}
# Remove the unclassified observations
ncsr_biota_clean <- ncsr_biota %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
ncsr_biota_simple <- ncsr_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
biota_ncsr <- ncsr_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /ncsr_area) * 100)
```

```{r}
 # Create table
kable(biota_ncsr, col.names = c("Biota Category", "Area in hectares", "Percent of Total North Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In NCSR MPAs

```{r}
# Read in the prefiltered biota data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
ncsr_biota <- readRDS(file.path(rds.dir, "ncsr_biota.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
norcal_biota_transform <- st_transform(ncsr_biota, crs = 32610)

# Calculate area in hectares
norcal_biota_transform$norcal_ha <- (st_area(norcal_biota_transform) / 10000)

# Remove the unclassified observations
norcal_biota_clean <- norcal_biota_transform %>%
  filter(cmecs_bc_category != "Unclassified")

# Drop geometries for ease of processing
norcal_biota_simple <- norcal_biota_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
norcal_biota_simple_2 <- norcal_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(norcal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_biota = (biota_area /ncsr_area ) * 100)
```

```{r}
 # Create table
kable(norcal_biota_simple_2, col.names = c("Biota Category", "Area in hectares", "Percent of Total North Coast Region MPAs")) %>%
  kable_material(c("striped", "hover"))
```



