---
title: "estuary_components"
author: "Maddy Enda"
format: html
editor: visual
---

```{r}
# Load in the packages for this data analysis
librarian::shelf(tidyverse, dplyr, terra, tmap, sf, here, stars, janitor, units, showtext)
```

```{r}
# Adding in our text
font_add_google("Montserrat", "montserrat")
showtext_auto()

# Create palette
biota_pal <- c("seagreen", "#A3AD62FF", "#D0D3A2", "#798234FF", "#F0C6C3FF", "#DF91A3FF", "#D46780FF" )
```

```{r}
# Unpack the biota layer and filter to only California observations
estuary <- sf::read_sf(here::here("data", "PMEP_West_Coast_USA_Estuary_Extent_V1.gdb")) %>%
  janitor::clean_names()
```

```{r}
# Load in the MPA boundary data
MPA_boundary <- sf::st_read("data/California_Marine_Protected_Areas_[ds582]")

# Make the colummn names lower snakecase
mpa <- MPA_boundary %>%
   clean_names()
```

```{r}
# Match the crs of the two data sets
estuary <- st_transform(estuary, crs= st_crs(mpa))
```

```{r}
# Join the data sets together
estuary_mpa <- st_intersection(estuary, mpa)
```

-   We get an error here that tells us the geometries are multisurface and not compatible with multipolygons.

```{r}
# Make the multisurface geometries of the estuary data set vaild sf objects
estuary_valid <- st_as_sf(estuary)

# Convert to a multipolygon type
estuary_polygon <-st_cast(estuary_valid, "MULTIPOLYGON")
```

```{r}
# Join the data sets together
estuary_mpa <- st_intersection(estuary_polygon, mpa)
```

-   Success! We have them matched and have taken the intersection

### Find percent of CA that is biota

```{r}
{r}
# Create a column that has the total area of socal biota
ca_biota_simple <- ca_biota_simple %>%
  filter(cmecs_bc_category != "Floating/Suspended Plants and Macroalgae") %>%
  mutate(total_area = sum(area_ha))

# Create a sum of area column in mpa_transform
ca_biota_simple_2 <- ca_biota_simple %>%
  group_by(cmecs_bc_category) %>%  # Group by MPA type, substrate group
  summarise(
    biota_area = sum(area_ha, na.rm = TRUE),  
    total_area = first(total_area)  
  ) %>%
  mutate(pct_biota = (biota_area / total_area) * 100)
```

### Find percent of MPAs that are biota

```{r}
head(estuary_mpa)
```

```{r}
# Calculate area once more with st_area
estuary_mpa$mpa_area_m <- st_area(estuary_mpa)
estuary_mpa <- estuary_mpa %>%
  mutate(mpa_area_ha = mpa_area_m / 1000)

# Drop the geometries to speed up run time
estuary_mpa_simple <- estuary_mpa %>%
  st_drop_geometry()
```

```{r}
# Create a column that has the total area of socal biota
estuary_mpa_simple <- estuary_mpa_simple %>%
  mutate(total_mpa_ha = sum(mpa_area_ha))

# Create a sum of area column in mpa_transform
estuary_mpa_simple_2 <- estuary_mpa_simple %>%
  group_by(cmecs_class) %>%  # Group by MPA type, substrate group
  summarise(
    estuary_area = sum(mpa_area_ha, na.rm = TRUE),  
    total_area = first(total_mpa_ha)  
  ) %>%
  mutate(pct_biota = (estuary_area / total_area) * 100)
```

```{r}
ggplot(estuary_mpa_simple_2, aes(x = pct_biota, y = reorder(cmecs_class, pct_biota), fill = cmecs_class)) +
  geom_col() +
  labs(
    title = " Estuary Type Distribution \n\ within the CA MPA Network",
    x= "Percent of Total Estuary Area",
    y = "PMEP Biotic Class"
  ) +
  scale_fill_manual(values = biota_pal) +
  theme_classic() +
  theme(text = element_text(family = "montserrat", size = 15),
        title = element_text(family= "montserrat", size= 13),
    legend.position = 'none',
    axis.text.x = element_text(hjust = 1)) +
  geom_label(aes(label = paste0(round(pct_biota, 1), "%")),
             fill = "white", 
             color = "black",
             hjust = 0,    # Align to left
             x = 1,        # Set near y-axis (adjust as needed)
             size = 3.5) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 20))
```

```{r}
tm_shape(estuary_polygon) +
  tm_polygons()
```


```{r}
```{r}
# Unpack the biota layer and filter to only California observations
estuary <- sf::read_sf(here::here("data", "PMEP_West_Coast_USA_Estuary_Extent_V1.gdb")) %>%
  janitor::clean_names()

# Project UTM 10 crs for calculating area
estuary <- st_transform(estuary, crs= st_crs(32610))

# Calculate area once more with st_area
estuary$mpa_area_m <- st_area(estuary)
estuary <- estuary %>%
  mutate(mpa_area_ha = mpa_area_m / 1000)

# Drop the geometries to speed up run time
ca_estuary_simple <- estuary %>%
  st_drop_geometry()

# Create a column that has the total area of socal biota
ca_estuary_simple <- ca_estuary_simple %>%
  mutate(total_ha = sum(mpa_area_ha))
```
```

